Kindly Bent to Free Us

### 2019-11-21 kindly/lang.tex
Misc.

In the examples presented in \cref{motivation}, region annotations are optional.
In our formalization, regions must be syntactically explicit and annotated with
a nesting index and a scoped variable.
We now define a best-effort
syntactic code transformation $\RannotT{p}{p'}$ which
annotates programs automatically.
The input $p$ is a program with optional region annotations of the form $\region[]{}{e}$.
The output $p'$ is a program with explicit annotations
of the form $\region{\Sone x\BORROW}{e}$.
We give an informal presentation of our code transformation. The complete
definition is given in \cref{appendix:regionannot}. This code transformation
aims to find, for each borrow $\borrow{x}$, the biggest region satisfying
the following rules:
\begin{enumerate}
\item The region should contain at least $\borrow{x}$.
\item The region should never be larger than the scope of $x$.
\item An exclusive borrow $\borrow[\kaff]{x}$ should never share a region with any other borrow of $x$.
\item A use of $x$ can never be in the region of $\borrow{x}$.
\end{enumerate}
Following these rules, we start from the leafs and grow the region
associated to each borrow until it can't grow anymore either due to a binding
or to conflicting uses.

Let us consider the following program with binders and function applications:
%
\[
  \setlength{\jot}{-1pt}
  \RannotT{
\begin{aligned}
  \lam{a}{}\ &
  \letin{x}{\app{f}{\borrow[\kun]{a}}}{}\\
  &g\ (\borrow[\kaff]{x});\\
  &f\ (\borrow[\kun]{x})\ (\borrow[\kun]{x})\\
\end{aligned}
}{
\begin{aligned}
  \lam{a}{}\region[1]{\Sone a\MBORROW}{\ &%
  \letin{x}{\app{f}{\borrow[\kaff]{a}}}{}\\
  &\region[2]{\Sone x\MBORROW}{g\ (\borrow[\kaff]{x})};\\
  &\region[2]{\Sone x\IBORROW}{f\ (\borrow[\kun]{x})\ (\borrow[\kun]{x})}\\
  &\!\!\!}
\end{aligned}
}
\]
%
Since the only use of $a$ is a borrow,
the region covers the entire lexical scope of $a$. For $x$ however, there are
multiple conflicting borrow and we must be more careful. We place a first
region around the exclusive borrow and its function application,
and a second region around both shared borrows. This placement of region
is optimal, as making any of the region bigger would cause a borrowing
error.
Notes that indices of regions are well nested, as the inner region have
higher indices that the outer ones.

@@

\subsubsection{Kinds and constraints}

@@

Kinds form a lattice with an ordering that implements two principles
(see \cref{sdtyp:lattice}). First, unrestricted values may be used
where affine ones are expected and affine ones are less restricted
than linear ones as reflected in \TirName{Lat-UAL}.
Second, values usable at level $n$ may be used at any more deeply
nested level $n'$ (where $n\le\infty$ for any $n$) as defined in the \TirName{Lat-Level} axioms.

Constraints are conjunctions of inequality constraints over this kind
lattice. In most uses, a constraint gives an upper or lower bound for
a kind variable or it relates two kind variables.

@@

@@

\subsubsection{Environments and bindings}
\label{sdtyping:envs}

\lang controls the use of variables by supporting new modes of
binding in a type environment $\E$.
Its definition is given in  \cref{grammar:env}.
In addition to standard bindings of type variables to kind schemes,
$\bvar{\tvar}{\kschm}$,  and value bindings, $\bvar{x}{\schm}$, there
are suspended bindings and borrow bindings.
A suspended binding, $\svar{x}{\schm}^n$, indicates that $x$ is
earmarked for a borrowing use in a nested region
marked with $x$ (see \cref{sdtyping:regions}), but
cannot be used directly.
A borrow binding, $\bvar{\borrow{x}}{\borrowty{k}{\schm}}$, replaces
such a suspended binding on entry to the $x$-region. It indicates
that the borrow $\borrow{x}$ can be used directly. The kind $k$
restricts the lifetime of the borrow to the region (see \cref{selectrules:borrow}).

@@

\subsubsection{Copying and Splitting}
\label{sdtyping:split}

@@

\subsubsection{Pattern matching}
\label{sdtyping:matching}

@@

\subsubsection{Regions}
\label{sdtyping:regions}

@@

% \subsubsection{Kind checking}

@@

\subsubsection{Resource management}

>>>>

In the examples presented in \cref{motivation}, region annotations are optional.
In our formalization, regions must be syntactically explicit and annotated with
a nesting index and a scoped variable.
We now define a best-effort
syntactic code transformation $\RannotT{p}{p'}$ which
annotates programs automatically.
The input $p$ is a program with optional region annotations of the form $\region[]{}{e}$.
The output $p'$ is a program with explicit annotations
of the form $\region{\Sone x\BORROW}{e}$.
We give an informal presentation of our code transformation. The complete
definition is given in \cref{appendix:regionannot}. This code transformation
aims to find, for each borrow $\borrow{x}$, the biggest region satisfying
the following rules:
\begin{enumerate}
\item The region should contain at least $\borrow{x}$.
\item The region should never be larger than the scope of $x$.
\item An exclusive borrow $\borrow[\kaff]{x}$ should never share a region with any other borrow of $x$.
\item A use of $x$ can never be in the region of $\borrow{x}$.
\end{enumerate}
Following these rules, we start from the leafs and grow the region
associated to each borrow until it can't grow anymore either due to a binding
or to conflicting uses.
%
Let us consider the following program with binders and function applications:
%
\[
  \setlength{\jot}{-1pt}
  \RannotT{
\begin{aligned}
  \lam{a}{}\ &
  \letin{x}{\app{f}{\borrow[\kun]{a}}}{}\\
  &g\ (\borrow[\kaff]{x});\\
  &f\ (\borrow[\kun]{x})\ (\borrow[\kun]{x})\\
\end{aligned}
}{
\begin{aligned}
  \lam{a}{}\region[1]{\Sone a\MBORROW}{\ &%
  \letin{x}{\app{f}{\borrow[\kaff]{a}}}{}\\
  &\region[2]{\Sone x\MBORROW}{g\ (\borrow[\kaff]{x})};\\
  &\region[2]{\Sone x\IBORROW}{f\ (\borrow[\kun]{x})\ (\borrow[\kun]{x})}\\
  &\!\!\!}
\end{aligned}
}
\]
%
Since the only use of $a$ is a borrow,
the region covers the entire lexical scope of $a$. For $x$ however, there are
multiple conflicting borrow and we must be more careful. We place a first
region around the exclusive borrow and its function application,
and a second region around both shared borrows. This placement of region
is optimal, as making any of the region bigger would cause a borrowing
error.
Notes that indices of regions are well nested, as the inner region have
higher indices that the outer ones.

@@

\paragraph{Kinds and constraints}

@@

Kinds form a lattice with an ordering implementing two principles
(see \cref{sdtyp:lattice}). Unrestricted values may be used
where affine ones are expected and affine ones are less restricted
than linear ones as reflected in \TirName{Lat-UAL}.
Values usable at level $n$ may be used at any more deeply
nested level $n'$ (where $n\le\infty$ for any $n$) as defined in the \TirName{Lat-Level} axioms.

Constraints are conjunctions of inequality constraints over this kind
lattice. In most uses, constraints give upper or lower bounds for
kind variables or relate two kind variables.

@@

\paragraph{Environments and bindings}
\label{sdtyping:envs}

@@

\lang controls the use of variables by supporting new modes of
binding in a type environment $\E$.
Its definition is given in  \cref{grammar:env}.
In addition to standard bindings of type variables to kind schemes,
$\bvar{\tvar}{\kschm}$,  and value bindings, $\bvar{x}{\schm}$, there
are suspended bindings and borrow bindings.
A suspended binding, $\svar{x}{\schm}^n$, indicates that $x$ is
earmarked for a borrowing use in a nested region
marked with $x$ (see \cref{sdtyping:regions}), but
cannot be used directly.
A borrow binding, $\bvar{\borrow{x}}{\borrowty{k}{\schm}}$, replaces
such a suspended binding on entry to the $x$-region. It indicates
that the borrow $\borrow{x}$ can be used directly. Kind $k$
restricts the lifetime of the borrow to the region (see \cref{selectrules:borrow}).

@@

\paragraph{Copying and Splitting}
\label{sdtyping:split}

@@

\paragraph{Pattern matching}
\label{sdtyping:matching}

@@

\paragraph{Regions}
\label{sdtyping:regions}

@@

% \paragraph{Kind checking}

@@

\paragraph{Resource management}

### 2019-11-21 kindly/motivation.tex
XXX Do something horrifying to whitespaces.
Misc.

After
creation of the array, the presence of a borrow in the for loop
prevents access to the ``raw'' resource inside the loop's
body. \cref{ex:array:get} of the example contains two shared borrows
in the same expression, which forms a region by itself. These borrows
are split off the exclusive borrow used in \cref{ex:array:set}, which
belongs to the next enclosing region corresponding to the loop body.
The whole array can be returned in \cref{ex:array:return} because  the
borrows are no longer in scope.

More precisely, the code below is an annotated excerpt with explicit regions
marked in braces @{| ... |}@:

@@

\begin{figure}[tp]
  \lstinputlisting[linerange=1-31]{code/sukodu.affe}
  \vspace{-15pt}
  \caption{Excerpt of the Sudoku solver}
  \label{ex:sudoku}
\end{figure}

On top of the mutation function @set_mut@, our array implementation
also provides @set@ of type @&('a t) * int * 'a -> 'a t@. It
takes a shared borrow because it only reads from the
argument array, but returns a fresh, modified structure.
%
This API could be implemented in a wide variety of ways, such as
copy-on-write (which copies the whole array on each @set_mut@)
or something more sophisticated\cite{DBLP:conf/ml/ConchonF07}.

>>>>

After
creation of the array, the presence of a borrow in the for loop
prevents access to the ``raw'' resource inside the loop's
body. \cref{ex:array:get} of the example contains two shared borrows
in the same expression, which forms a region by itself. These borrows
are split off the exclusive borrow used in \cref{ex:array:set}, which
belongs to the next enclosing region corresponding to the loop body.
The whole array can be returned in \cref{ex:array:return} because  the
borrows are no longer in scope. 
%
More precisely, here is is an annotated excerpt with explicit regions
marked in braces @{| ... |}@:

@@

\begin{figure}[tp]
  \lstinputlisting[linerange=1-31]{code/sukodu.affe}
  \vspace{-15pt}
  \caption{Excerpt of the Sudoku solver}
  \label{ex:sudoku}
  \vspace{-10pt}
\end{figure}

On top of the mutation function @set_mut@, our array implementation
also provides @set: &('a t) * int * 'a -> 'a t@. It
takes a shared borrow because it only reads from the
argument array, but returns a fresh, modified structure.
%
This API could be implemented in a wide variety of ways, such as
copy-on-write (which copies the whole array on each @set_mut@)
or something more sophisticated\cite{DBLP:conf/ml/ConchonF07}.
